Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-03 method.tex
merged
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@908 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the compact binary parameter
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in a
search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 science runs.
We combine this with the orthogonal decomposition described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Inspiral signals are continuously parameterized by a set of intrinsic source parameters
$\vec\theta$ that determine the amplitude and phase evolution of the gravitational wave strain. For systems where the effects of spin can be ignored, the
intrinsic source parameters are the component masses of the binary, $\vec\theta =
(m_1, m_2)$. Searches for inspiral signals
typically employ matched filter banks that discretely
sample the possible intrinsic parameters~\cite{findchirppaper}.  For a given source, the strain observed by the detector is a linear combination of two waveforms corresponding to the `$+$' and `$\times$' gravitational wave polarizations, so for any value of $\vec\theta$ we must implement 2 filters.  The coefficients for the $M$ filters
are known as templates, and are formed by discretizing and time reversing the waveforms and weighting them by the inverse amplitude spectral density of the detector's noise. To construct a template bank, templates are chosen with the $M/2$ discrete signal parameters $\vec\theta_0,\, \vec\theta_1,\, \dots,\, \vec\theta_{M/2-1}$ to assure a
bounded loss of \SNR~\cite{Owen:1995tm,Owen:1998dk}\editorial{Can we cut this down to one citation?}. That is, any possible
signal within the a given range of intrinsic source parameters will have an inner product that is $\geqslant 0.97$ with
at least one template. Such a template bank is said to have a {\em
minimum match} of 0.97. The data from the detector are whitened and convolved with each template to produce $2M$ \SNR\
time series. Local peak-finding across time and templates determines detection
candidates.

@@

Equation~(\ref{eq:SNRTD}) can be implemented in the time domain as an \fir{} filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point operations per sample.  However, it is typically much more computationally efficient to use the convolution theorem and the \fft{} to implement fast convolution, requiring only $\mathcal O(\numtmps \lg \tmpsamps)$ operations per sample.

@@

In the remainder of this section we explore a method for reducing the
computational cost of a time-domain search for compact binary coalescence.  We
will give a truly zero-latency algorithm that competes in terms of floating point
operations per second with the conventional frequency-domain method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of orthogonal
filters with far fewer coefficients than the original templates.  The reduction
in the number of sample points in all of the filters required to search the entire
parameter space is dramatic.

The first transformation is to chop the time-domain templates up into time
slices.  Since each template piece is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim 10$ samples on
%
\editorial{Where do these numbers come from?}
%
average and also allows the filters to be evaluated at about a $\sim 10$ times
lower rate on average.  This amounts to more than a factor of 100 in the
floating point operations per second required.  However, the resulting filters
are still not orthogonal across the mass parameter space, and are in fact
highly redundant.  We use the singular value decomposition to produce an
orthogonal filter set from the time sliced templates~\cite{Cannon:2010p10398}.
We find that this reduces the number of sample points that need to be filtered
by another factor of $\sim 100$.  The combined methods reduce the number of
floating point operations to the level where they are competitive with the
conventional matched filter approach.  In the remainder of this section we
describe the \lloid\ algorithm in detail and provide some basic computational
cost scaling.

@@

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency $f_\mathrm{hi}$.
This is often chosen to correspond to the \ISCO. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t^0)$ and is said to have  a
duration of $t^0 - t_\mathrm{low}$. It is critically sampled at a
rate of $2 f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the convolution at a lower sample
rate.  Specifically we consider here time slice boundaries with the smallest
power-of-two sample rate that sub-critically samples the time sliced
template.  The time slices consist of the $S$ intervals $\left(t^S,
t^{S-1}\right],\, \dots,\, \left(t^2, t^1\right],\, \left(t^1, t^0\right]$ sampled at frequencies
$f^\mathrm{S-1},\, \dots,\, f^1,\, f^0$ where $f^0 \geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time-sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of orthonormal {\em
basis templates}.  A numerical technique based on the
singular value decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described above
can be approximated to arbitrary accuracy by expansion into a set of orthonormal
\emph{basis templates}, $u_l^s[k]$, related to the original time sliced templates
through the \emph{reconstruction matrix}, $v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ sets the number of basis templates that are kept in the
approximation.  This determines the \SVD\ tolerance, which affects the \SNR\ loss due
to the approximation.  The authors of \cite{Cannon:2010p10398} showed that high accuracy
could be achieved with far fewer basis templates than templates in the original template
bank.  We find that when combined with the time slice decomposition, the number of basis 
templates \numsvdtmps\ is much smaller than the original number of templates \numtmps.
In the next section we describe how we form our novel detection statistic using the time slice decomposition and the \SVD.

\subsection{Proposed method}

Definition of decimated detector data:
\begin{equation}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k]
\end{equation}

\editorial{Chad: This whole subsection needs a lot of work.
I just dropped our new notation in here.}%
In order to minimize latency we propose using the time-domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute-force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational waves from compact binary coalescence in \realtime.  This
requires 1) exploiting
the time-frequency characteristics of the binary waveforms and 2) exploiting the redundancy of the template bank. We describe our
procedure for producing the decomposition in \eqref{eq:decomp} in the remainder
of this section.

@@

>>>>

In this section we describe a decomposition of the compact binary parameter
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR2 science runs.
We combine this with the \SVD\ rank reduction method described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Inspiral signals are continuously parameterized by a set of intrinsic source
parameters $\vec\theta$ that determine the amplitude and phase evolution of the
gravitational wave strain. For systems where the effects of spin can be
ignored, the intrinsic source parameters are the component masses of the
binary, $\vec\theta = (m_1, m_2)$. Searches for inspiral signals typically
employ matched filter banks that discretely sample the possible intrinsic
parameters~\cite{findchirppaper}.  For a given source, the strain observed by
the detector is a linear combination of two waveforms corresponding to the
`$+$' and `$\times$' gravitational wave polarizations, so for any value of
$\vec\theta$ we must implement 2 filters.  The coefficients for the $M$ filters
are known as templates, and are formed by discretizing and time reversing the
waveforms and weighting them by the inverse amplitude spectral density of the
detector's noise. To construct a template bank, templates are chosen with the
$M/2$ discrete signal parameters $\vec\theta_0,\, \vec\theta_1,\, \dots,\,
\vec\theta_{M/2-1}$ to assure a bounded loss of
\SNR~\cite{Owen:1995tm,Owen:1998dk}\editorial{Can we cut this down to one
citation?}. That is, any possible signal within a given range of intrinsic
source parameters will have an inner product that is $\geqslant 0.97$ with at
least one template. Such a template bank is said to have a {\em minimum match}
of 0.97. The data from the detector are whitened and convolved with each
template to produce $2M$ \SNR\ time series. Local peak-finding across time and
templates determines detection candidates.

@@

Equation~(\ref{eq:SNRTD}) can be implemented in the time domain (\TD) as an
\fir\ filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point
operations per sample.  However, it is typically much more computationally
efficient to use the convolution theorem and the \fft\ to implement fast
convolution in the frequency domain (\FD), requiring only $\mathcal O(\numtmps
\lg \tmpsamps)$ operations per sample.

@@

In the remainder of this section we explore a method for reducing the
computational cost of a \TD\ search for compact binary coalescence.  We will
give a truly zero latency algorithm that competes in terms of floating point
operations per second with the conventional \FD\ method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.
The reduction in sample count for all of the filters required to search the
entire parameter space is dramatic.

The first transformation is to chop the time-domain templates up into time
slices.  Since each template slice is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim 100$ samples
from
%
\editorial{Where do these numbers come from? CHAD: I have actually changed the statement to 100 for the early time slices.}
%
the early part of the waveform and allows the filters to be evaluated at about
$\sim 100$ times lower rate.  This amounts to more than a factor of 10000
reduction in the floating point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of $\sim
100$.  The combined methods reduce the number of floating point operations to
the level where they are competitive with the conventional high latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency
$f_\textrm{hi}$.  This is often chosen to correspond to the \ISCO. An inspiral
signal will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t^0)$ and is said to have  a duration of
$t^0 - t_\mathrm{low}$. It is critically sampled at a rate of $2
f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.  Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time sliced
template.  The time slices consist of the $S$ intervals $\left(t^S,
t^{S-1}\right],\, \dots,\, \left(t^2, t^1\right],\, \left(t^1, t^0\right]$
sampled at frequencies $f^\mathrm{S-1},\, \dots,\, f^1,\, f^0$ where $f^0
\geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time
sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}.  A numerical technique based on the singular value
decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ sets the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps.  In the next section we
describe how we form our novel detection statistic using the time slice
decomposition and the \SVD.

\subsection{Early warning \SNR }

In the previous two sections we have described two transformation that greatly
reduce the burden of filtering the compact binary parameter space and make \TD\
filtering of the data possible.  We are now prepared to define the early
warning \SNR\ and to comment on the computational cost of evaluating it.  But
first, we will introduce some notation refering to the decimation of data,
which means reducing the sample rate after low pass filtering,
\begin{equation}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k]
\end{equation}
and notation refering to the interpolation of data to increase the sample rate,
\begin{equation}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k]
\end{equation}

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
This quantity is related to the early warning \SNR\ by the normalization of
the partial filter defined by the present time slice.  This normalization
may be computed in the planning stage and can be easily multiplied here.
Figure \ref{fig:pipeline} describes the computation of the \SNR\ and
early warning \SNR\ schematically.
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of LLOID pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for FIR filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram drawn in figure \ref{fig:pipeline} contains several
distinct stages and parallel branches.  The stages are decimation, filtering
the data against the orthogonal templates, reconstruction of the original
template basis, interpolation and accumulation of the early warning and final
{\SNR}.

@@

\subsubsection{Extrapolation of the computational cost of the \lloid\ method to an Advanced LIGO search}

Table \ref{table:flops} shows that it requires about 0.5 GFLOPS to conduct the
filtering, reconstruction and resampling stages of the \lloid\ method for 657
templates chosen around typical NS--NS mass parameters.  Given the capability
of modern CPUs it should be possible to filter that sub-bank on one CPU.
Assuming that the cost holds for the other sub-banks we expect that the full
parameter space should require $\sim 150$ CPUs for realtime filtering with the
\lloid\ method per detector.  Even if real implementations fall short of these
estimates by factors of several, there will still be ample computing resources
for an early warning search in the advanced detector era.
